### 2021-08-30 response_to_review.Rmd

- p3 l28 you mention Andrews curves later on when you speak about software; mention them here as well and clarify that they are some kind of tour, but not a grand tour; Wegman & Solka (2002) describe Andrews plot as a pseudo-grand tour that is not space filling - see
On Some Mathematics for Visualizing High Dimensional Data
Author(s): Edward J. Wegman and Jeffrey L. Solka
Source: Sankhyā: The Indian Journal of Statistics, Series A (1961-2002) , Jun., 2002, Vol.
64, No. 2, Selected Articles from San Antonio Conference in Honour of C. R. Rao (Jun.,
2002), pp. 429-452
Published by: Indian Statistical Institute
Stable URL: https://www.jstor.org/stable/25051404

    - "Added, and explained"

- p3 l28 Wegman & Solka (2002) also provide mathematical details and further references; while it is not necessary for this article to repeat all of that information, it should at least summarize those mathematical references in one paragraph.

    - "Not done. We have reference the paper but the Buja et al is the better reference for mathematics."

@@

- p4 l41 add a "," here

    - **Done**

- p5 fig 3 what does flat and hierarchical mean here?

    - **removed legend because these labels are not really sensible or useful**

@@

- p15 l11 add "that"

- p15 l21 indicate the names of the software, i.e., XplorN and (X)Lisp-Stat; in particular, XplorN deserves some further description as it was one of the first software implementations that provided a PCP grand tour; here is a suitable technical report (follow the cite references for more):
https://apps.dtic.mil/sti/pdfs/ADA313999.pdf

- p15 l22 this paragraph doesn't seem to belong here as it doesn't seem to relate to software; I suggested earlier on that Andrews curves should be introduced earlier in this review; this paragraph can be used there

    - **Moved to the intro with other discussion of lietrature**

@@

- p19 l19 add ","

- p19 l19 be specific what is missing in current software, i.e., continue with "(such as ...)."

- p19 l23 perhaps "spaces" (there could be more than just one?)

- References
    - video library url needed - done
    - buja et al in wegman & solka add city - done
    - buja et al (1986) this is a conference proceedings series; capitalize similar to a journal - I think this needs to be fixed via bibstyle
    - carr add page numbers (or is this the previous ref) - I think this is for the previous reference, again it is there in the bib file, to show it the bibstyle should be formatted accordingly for proceeding entries
    - chambers et al list the editors - I don't think there was any 'scientific' editor on this book, didn't find any relevant person to list here ...
    - coleman add when accessed - this was taken from animation package so I got the date when the rda file was added to its github repo and added it here...
    - cook et al (1995) adjust style (compare with previous reference) - done
    - cook et al (2018) remove the ArXiv information for an article that has been published - done
    - cook, swayne (2007) reference is wrong add add city of publication (New York, Berlin, Heidelberg?) - done
    - diaconis as above - either full journal names or abbreviations (adjust other references in a similar way, according to WIREs specifications) - done
    - fisher remove the first URL and just keep the one with the DOI - done
    - friedman is there a DOI? - added
    - horst capitalize both "A" as these seem to be geographic names; just "R package" once - fixed capitalization, but repetition needs to be fixed in bibstyle
    - hotelling should be "417-441" - fixed (plus full journal name instead of abbreviation)
    - huber is there a URL where this can be obtained? - could not find it via google at least...
    - laa et al (2020) section pursuit incomplete - fixed
    - laa et al (2021) burning sage incomplete - fixed
    - laa et al (2020) slice tour update reference (it is published now) and remove the arXiv information - fixed
    - lee at al DOI ? - added
    - li incomplete; use lower case "c" and "t" here - fixed
    - li where does the 5(3) come from? - check the web page again - fixed
    - martinez remove the first URL - fixed
    - mcinnes capitalize "UMAP" - fixed
    - moustafa incomplete; Remove the first URL - fixed
    - moustafa and hadi remove the first URL - fixed
    - nguyen capitalize P (in Python), R (in the next reference), etc. - fixed
    - pedersen just once; check gganimate capitalisation - fixed capitalisation, but repeating version number is bibstyle issue
    - sievert city?; should be "Chapman and Hall/CRC" (several times) - fixed
    - spyrison change to lower case (except R); 243-257 (again!) - fixed
    - sutherland et al DOI - added
    - swayne DOI - I have no idea why it is not printing the DOI for the XGobi reference...
    - swayne et al DOI; XGobi (as a software name); issue ? -fixed
    - tukey editors? - added something based on web search, but not sure I got the right names..
    - tukey pages, city, publisher ?; city? - as above ..
    - van der maaten they list this as 9(86) on their web page - fixed
    - wegman list editors only once; page numbers; is there a volume ? - these are Interface (conference proceedings) - problem with showing these information for proceedings entry, again bibstyle issue?
    - wilkinson capital "S" - fixed
    - xie lower case - fixed
    - you "R" - fixed
    - zhang extend reference as for other R packages; incomplete - fixed

### 2021-08-31 paper.Rmd
fixing tsne plot and cleaning up intro

Data commonly arrives with more than two measured variables, which makes it more complicated to plot on a page. With multiple variables, especially if there is some association between variables, this would be called multivariate or high-dimensional data. When the variables are all numeric, or quantitative, visualization often relies on some form of dimension reduction. This can be done by taking linear projections, for example, principal component analysis [@Hotelling1933-of] or linear discriminant analysis [@Fisher1936]. It is also common to reduce dimension with nonlinear techniques like multidimensional scaling (MDS) [@Kruskal1964-do] or t-Distributed Stochastic Neighbor Embedding (t-SNE) [@Van_der_Maaten2008-qa].

@@

The focus of this review will be on visualizing high-dimensional numerical data using linear projections, in particular, as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. The reason being that it is not feasible to adequately review the very large area of visualizing high-dimensions, and there have been numerous developments in tours recently. An overview of the technique, and new modifications is provided, along with how these techniques can be used in a variety of applications.

A tour can be considered to be a dynamic graphic, because it shows a smooth sequence of projections over time, ideally with controls that allow stopping, reversing, changing direction, or going forward again. It can be useful to embed a tour into an interactive graphics system, where plots can be queried and elements highlighted (e.g. GGobi @Swayne2003 or XLispStat @Tierney1991). To create the smooth sequence, a geodesic interpolation is computed between consecutive frames. A geodesic interpolation produces movement along the shortest path between two planes. It allows the viewer to extrapolate from the low-dimensional shapes to a corresponding multivariate distribution, and is particularly useful for detecting clusters, outliers and non-linear dependence.

@@

Nonlinear dimension reduction techniques such as t-SNE and Uniform Manifold Alignment and Projection (UMAP) have become very popular in recent years, primarily for the ability to capture cluster structure in a succinct visual summary [@Van_der_Maaten2008-qa;@mcinnes2020umap]. However, it is only a summary, and it likely involves substantial warping of the original data space. Using the tour along with these techniques can illuminate the nature of the warped space, and reveal other structure lost in the dimension reduction.  Figure \ref{fig:tsne-tour} illustrates the difference in what can be learned from a tour in comparison with nonlinear dimension reduction using t-SNE. The 10-dimensional data comes from @Rauber2009-kk. The t-SNE view (plot A) shows six clearly separated clusters, all spherical, which have different sizes. The tour shows that the clusters do not look like this in the full data space. The clusters are at various distances apart and very different sizes, as can be seen from the four projections from a tour. The two green clusters are large and almost spherical, and far from the orange clusters. The orange clusters have one larger one, and three smaller, very close to each other. All of these are elliptical, which means that they have very little variability, actually no variability if you watch the full tour, in some of the 10 dimensions. This gives some deeper perspective to what is learned from t-SNE, and illustrates what the t-SNE dimension reduction has done: it has found small gaps between points and expanded these gaps to yield the representation. It should be noted, though, that the methods (t-SNE and tour) complement each other. The t-SNE view gives a clear indication of six clusters, which may have been overlooked on initial viewing with a tour. With this information is the invitation to look closer at the data in the tour, to see, that yes, indeed, there are three, tiny, tiny clusters very close to each other.

\begin{figure*}[!t]
         \includegraphics[width=0.255\textwidth]{figures/cluster_example.pdf}
     \vfill
   \begin{minipage}{\linewidth}     
   \centering
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour1.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour2.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour3.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour4.png}
   \end{minipage}%
\caption{Comparison of structure perception between nonlinear dimension reduction from 10D, using t-SNE (A) and a tour (B). There are six clusters, visible in the t-SNE view, but the relative distance between the clusters is extremely varied. This can be seen in a selection of tour projections shown. The two green clusters are (almost) spherical in shape, and very distant from the orange clusters. Three of the orange clusters are very close to each other (just visible in B.1), and all orange clusters are elliptical. The tour provides a more accurate rendering of the clusters in the high-dimensional space, and complements what is learned from the dimension reduction.}
\label{fig:tsne-tour}
\end{figure*}

@@

When using a tour, a sequence of $d$-dimensional linear projections are obtained from a $p$-dimensional space, where $d \ll p$. Let $X_{n \times p}$ be the data matrix, consisting of $n$ observations and $p$ variables, whose projections are of interest. A projection basis, $A_{p \times d}$, is matrix that characterizes the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires that each column in $A$ has unit length and are perpendicular. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows two examples of low-dimensional projections of the palmer penguins data [@penguins] in a Huber plot [@huber1990] and a histogram.

@@

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/notation-huber1.pdf}
         \includegraphics[width=\textwidth]{figures/notation-huber2.pdf}
\caption{Huber's plot (left plots) of the penguins data (bill length and depth only). The solid line represents how well the projection onto different directions, separates the three species, and the dashed circle is a reference guide. The histogram (right plots) is the 1D projection of the data onto the direction outlined as the solid line in the Huber's plot. Adapted from \cite{Lee2005-gl}, Figure 1.}
\label{fig:notation-huber}
\end{figure*}

@@

## Finding targets

>>>>

Data commonly arrives with more than two measured variables, which makes it more complicated to plot on a page. With multiple variables, especially if there is some association between variables, this would be called multivariate or high-dimensional data. When the variables are all numeric, or quantitative, visualization often relies on some form of dimension reduction. This can be done by taking linear projections, for example, principal component analysis (PCA) [@Hotelling1933-of], linear discriminant analysis (LDA) [@Fisher1936] or projection pursuit (PP) [@Friedman1974-ck]. It is also common to reduce dimension with nonlinear techniques like multidimensional scaling (MDS) [@Kruskal1964-do],  t-Distributed Stochastic Neighbor Embedding (t-SNE) [@Van_der_Maaten2008-qa] or Uniform Manifold Alignment and Projection (UMAP) [@mcinnes2020umap].

@@

The focus of this review will be on visualizing high-dimensional numerical data using linear projections, in particular, as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx] and its recent derivatives. The reason being that it is not feasible to adequately review the very large area of visualizing high-dimensions, and there have been numerous developments in tours recently. An overview of the technique, and new modifications is provided, along with how these techniques can be used in a variety of applications.

A tour can be considered to be a dynamic graphic, because it shows a smooth sequence of projections over time, ideally with controls that allow stopping, reversing, changing direction, or going forward again. It can be useful to embed a tour into an interactive graphics system, where plots can be queried and elements highlighted e.g. GGobi [@Swayne2003] or XLispStat [@Tierney1991]. To create the smooth sequence, a geodesic interpolation is computed between consecutive frames. A geodesic interpolation produces movement along the shortest path between two planes. It allows the viewer to extrapolate from the low-dimensional shapes to a corresponding multivariate distribution, and is particularly useful for detecting clusters, outliers and non-linear dependence.

@@

<!--Tours can be considered to be a dimension reduction technique. Visualizing high-dimensions often starts with dimension reduction, for example, principal component analysis (PCA) [@pearson-pca;@hotelling-pca;@joliffe-cadima-pca], which is a linear dimension reduction technique. PCA finds the linear combination (without the normality constraint of a projection vector) of maximum variance, and can be considered to be a special case of projection pursuit (PP) [@Friedman1974-ck] which maximizes an arbitrary function.-->

Nonlinear dimension reduction with the techniques t-SNE and UMAP, have become very popular in recent years, primarily for the ability to capture cluster structure in a succinct visual summary. However, it is only a summary, and it likely involves substantial warping of the original data space. Using the tour along with these techniques can illuminate the nature of the warped space, and reveal other structure lost in the dimension reduction.  Figure \ref{fig:tsne-tour} illustrates the difference in what can be learned from a tour in comparison with nonlinear dimension reduction using t-SNE. The 10-dimensional data comes from @Rauber2009-kk. The t-SNE view (plot A) shows six clearly separated clusters, all spherical, which have different sizes. The tour shows that the clusters do not look like this in the full data space. The clusters are at various distances apart and very different sizes, as can be seen from the four projections from a tour. The two green clusters are large and almost spherical, and far from the orange clusters. The orange clusters have one larger one, and three smaller, very close to each other. All of these are elliptical, which means that they have very little variability, actually no variability if you watch the full tour, in some of the 10 dimensions. This gives some deeper perspective to what is learned from t-SNE, and illustrates what the t-SNE dimension reduction has done: it has found small gaps between points and expanded these gaps to yield the representation. It should be noted, though, that the methods (t-SNE and tour) complement each other. The t-SNE view gives a clear indication of six clusters, which may have been overlooked on initial viewing with a tour. With this information is the invitation to look closer at the data in the tour, to see, that yes, indeed, there are three, tiny, tiny clusters very close to each other.

\begin{figure*}[!t]
   \centering
         \includegraphics[width=0.29\textwidth]{figures/cluster_example.pdf}
         \includegraphics[width=0.29\textwidth]{figures/cluster_example_zoom2/1.jpg}
         \includegraphics[width=0.29\textwidth]{figures/cluster_example_zoom2/5.jpg}
         \includegraphics[width=0.29\textwidth]{figures/cluster_example_zoom2/3.jpg}
         \includegraphics[width=0.29\textwidth]{figures/cluster_example_zoom3/1.jpg}
         \includegraphics[width=0.29\textwidth]{figures/cluster_example_zoom3/2.jpg}
     %\vfill
   %\begin{minipage}{\linewidth}     
   %\centering
   %      \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour1.png}
   %      \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour2.png}
  %       \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour3.png}
   %      \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour4.png}
  % \end{minipage}%
\caption{Comparison of structure perception between nonlinear dimension reduction using t-SNE (A) and linear dimension reduction using a tour (B) from 10D simulated data. There are six clusters visible in the t-SNE view, three large and three small, roughly equidistant. The relative distance between the clusters in the original space is extremely varied, as can be seen from the selection of tour projections shown. The two green clusters are (almost) spherical in shape, because they are spherical in every projection viewed. They are also far from all four orange clusters. Three of the orange clusters are tiny, and very close to each other (just visible in B.1, and clearer when zoomed in B.5). All orange clusters are elliptical. The tour provides a more accurate rendering of the clusters in the high-dimensional space, and complements what is learned from the nonlinear dimension reduction.}
\label{fig:tsne-tour}
\end{figure*}

@@

When using a tour, a sequence of $d$-dimensional linear projections are obtained from a $p$-dimensional space, where $d \ll p$. Let $X_{n \times p}$ be the data matrix, consisting of $n$ observations and $p$ variables, whose projections are of interest. A projection basis, $A_{p \times d}$, is a matrix that characterizes the direction from which the data are projected.  It needs to satisfy an orthonormality condition, which requires that all the columns in $A$ has unit length and are perpendicular to each other. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. The Huber plots and selected 1D projections shown in Figure \ref{fig:notation-huber} are a nice way to imagine the tour infrastructure. Two variables and the species labels from the palmer penguins data [@penguins] are used for illustration. In the Huber plots (left) a scatterplot of the two variables coloured by species is displayed in the center, and the solid line represents the value of illustrates the projection shows two examples of low-dimensional projections of the palmer penguins data [@penguins] in a Huber plot [@huber1990] and a histogram. (The figures produced with the R package PPtreeViz [@Lee2018-gy], and are motivated by @Lee2005-gl Figure 1.)

@@

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/notation-huber1.pdf}
         \includegraphics[width=\textwidth]{figures/notation-huber2.pdf}
\caption{Huber's plot (left plots) of the penguins data (bill length and depth only). The solid line represents how well the projection onto different directions, separates the three species, and the dashed circle is a reference guide. The histogram (right plots) is the 1D projection of the data onto the direction outlined as the solid line in the Huber's plot. These plots are made using the R package PPtreeViz.}
\label{fig:notation-huber}
\end{figure*}

@@

## Different target plane methods produce different tour types
